### 2020-07-16 manuscript/manuscript.rmd
add Dunn2008_choanozoa C60 results that previous not added
update manuscript with Nathan's comments

Yuanning Li: 0000-0002-2206-5804
Xing-Xing Shen: 0000-0002-8436-595X
Benjamin Evans:
Steven H.D. Haddock: 0000-0001-9420-4482
Antonis Rokas: 0000-0002-7248-6551
Casey W. Dunn: 0000-0003-0628-5150

@@

The models in the studies considered here differ according to two primary components: the exchangeability matrix $E$ and amino acid equilibrium frequencies $\Pi$. The exchangeability matrix $E$ describes the relative rates at which one amino acid changes to another. The studies considered here use exchangabilities that are the same between all amino acids (Poisson, also referred to as F81), fixed according to standard empirically estimated rates (WAG or LG), or all independently estimated from the data (GTR). The analyses considered here have site homogenous exchangeability models (site-homogeneous model), with the same matrix used for all sites in all genes. The equilibrium frequencies describe the expected frequency of each amino acid, which captures the fact that some amino acids are much more common than others. The analyses differ in whether they take a homogeneous approach and jointly estimate the same frequency across all sites, or add parameters that allow heterogeneous equilibrium frequencies that differ across sites. Heterogeneous approaches include CAT [@Lartillot:2004dq], which is implemented in the software PhyloBayes and has been widely applied to questions of deep animal phylogeny. The models that are applied in practice are heavily influenced by computational costs, model availability in software, and convention. For example, on the questions considered here, CAT site-heterogeneous models of equilibrium frequency have only been used in combination with Poisson and GTR exchangeability matrices, but not LG and WAG exchangeability matrices. Model effect is further confused by the abbreviations that are used for models. studies often discuss CAT and WAG models as if they are exclusive, but these particular terms apply to non-exclusive model components-- CAT refers to heterogeneous equilibrium frequencies across sites and WAG to a particular exchangeability matrix. In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. To avoid confusion on this point, here we always specify the exchangeability matrix first (e.g., GTR), followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (e.g., CAT). If site homogeneous equilibrium frequencies are used, we refer to the exchangeability matrix alone. Gamma-rate heterogeneity, a scalar that accommodates total rate of change across sites, is used in almost every analysis conducted here and we generally omit its designation.

High-throughput sequencing allows investigators to readily assemble matrices with hundreds or thousands of protein-coding genes from a broad diversity of animal species (Extended Data Table 1). Studies of animal phylogeny have used a wide variety of different approaches to identifying and selecting genes and taxa for their matrices. As a result, the particular genes selected for analysis vary widely (Fig. 2B, Extended Data Fig. 1). The composition of these genes varies in several ways, including the fraction of BUSCO and ribosomal protein genes in the matrix (Extended Data Fig. 2).

@@

**Fig. 2.** An overview of previous phylogenomic studies. (A) A time-series showing the main conclusions of previous phylogenomic studies that indicated their analyses supported Ctenophora-sister (orange nodes) or Porifera-sister (green nodes) based on their main conclusion from each study. (B) Each of the primary matrices considered here, color coded by taxon sampling. Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.

@@

**Fig. 3.**  Summary of phylogenomic results from previous studies and reanalysis conducted in this study. Each point represents a phylogenetic analyses with a support of bootstrap values less than `r bootstrap_threshold`% or a posterior probability less than `r posterior_prob_threshold`% are considered as unresolved. The clades are organized with increased outgroup sampling higher on the vertical axis, and the models are organized in general by increasing complexity in terms of numbers of parametersto the right on the horizontal axis. (A) Analyses transcribed from the literature (related to Supplementary Table 2). (B) New phylogenomic analyses conducted in this study (related to Supplementary Table 3).

@@

Our summary of previous phylogenetic analyses (Fig. 3A) show that Ctenophora-sister is supported in analyses that span the full range of outgroup sampling and models used to date, including some analyses with restricted outgroup sampling and models that accommodate site-heterogeneous equilibrium frequencies with CAT. This is consistent with previous assessments of the problem [@whelan2016let], but is drawn from a much more extensive and systematic examination. The only analyses that support Porifera-sister have a reduced outgroup sampling (Choanimalia, Holozoa) and site heterogeneous models with CAT. Model adequacy assessments generally favor GTR+CAT over Poisson+CAT or site-homogeneous models [@pisani2015genomic; @feuda2017improved], but because GTR+CAT is so parameter rich, many analyses that use a model with GTR+CAT do not converge and thus renders the analysis untrustworthy / hard to interpret. The fact that Porifera-sister is recovered only for particular models with particular outgroup sampling indicates that model and outgroup interact, and that this interaction is fundamental to understanding the range of results obtained across analyses.

@@

One of the challenges of interpreting support for the placement of the animal root across published studies is that different programs, software versions, and settings have been used across analyses. This extensive variation makes it difficult to identify the primary factors that lead to different results. Here we first reanalyze the primary matrices from each study under consistent conditions with IQ-TREE [@nguyen2014iq] with multiple evolutionary models. We selected this tool because it has greater model flexibility than most other phylogenetic tools [@zhou2017evaluating] (Fig. 3B; Supplementary Table 3). Importantly, it has C models (C10-C60 equilibrium frequencies) [@si2008empirical] that, like CAT, allow models to accommodate heterogeneity in equilibrium amino acid frequencies across sites.

@@

Among 15 previous studies,  three studies [@pisani2015genomic; @whelan2016let; @feuda2017improved]are based entirely on previously published data and gene-partition data is not available from one study [@Pick:2010eb]. For each of the published studies, we selected the matrix that was the primary focus of the manuscript for further analysis. For each of these matrices, we progressively trimmed taxon sampling to create Opisthokonta, Holozoa, and Choanimalia versions, where permitted by original outgroup sampling. This produced `r analyses_new$matrix%>% unique()%>% length()`data matrices from 11 studies.

@@

Once we established this large difference in category number, we sought to test its direct impact on animal root position. It is not possible to use more than 60 categories for C models in IQ-TREE, but the number of categories can be set *a priori* in Phylobayes using nCAT. We therefore varied the number of categories in Phylobayes analyses (Fig. 4; Supplementary Table 6). We found that Poisson+nCAT60 analyses in PhyloBayes, like Poisson+C60 and WAG+C60 IQ-TREE analyses, provide strong support of Ctenophora-sister. This indicates that the difference in results between unconstrained CAT analyses in PhyloBayes and C60 analyses in IQ-TREE is not due to differences in the software, but due to the large difference, in excess of ten-fold, in the number of site categories. When we varied the number of categories in PhyloBayes, we observe the transition from support for Ctenophora-sister to an unresolved root to Porifera-sister (Fig. 4). For example, for the Whelan2017_strict matrix this transition occurs between 60 -- 120 categories.

@@

These results further clarify when analyses support Porifera-sister. Outgroup sampling must be restricted (Choanimalia), a very large number of site categories must be used (unconstrained CAT, giving hundreds of categories), and a Poisson rather than GTR exchange matrix must be used. Analyses under other conditions either support Ctenophora-sister or are unresolved.

@@

ggarrange(ggarrange(p1, p2, ncol = 2, labels = c("A", "B")), nrow = 3,
          p3, p4, labels = c("C", "D"))

@@

Although a higher proportion of genes support Ctenophora-sister with site-heterogeneous C60 models, the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models. In an extreme case, in matrices from Ryan2013_est nearly 30% of genes changed from strong to week Ctenophora-sister signal ($\Delta$|lnL|<2) (Extended Data Fig. 3; Supplementary Table 7). In contrast to the C60 models, the phylogenetic signal largely increased in Poisson+CAT models in PhyloBayes, and the outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig. 3). For example in Whelan2017_full matrix, we found that the number of genes that support Ctenophora-sister in analyses with CAT decreases from 57.5% in matrices with distant outgroups (Holozoa) to 35.4 when outgroups are restricted (Choanimalia, Supplementary Table 7).

@@

Other approaches have been taken to addressing heterogeneity in equilibrium frequencies. For example, Feuda *et al.* [@Feuda:2017ew] recoded the full set of twenty amino acids into six groups of amino acids. These groups tend to have more frequent evolutionary changes within them than between them [@Susko:2007ds]. Recoding could, like CAT and C models, address variation across sites, but it could also accommodate variation across lineages, and it was suggested that this approach favors Porifera-sister [@Feuda:2017ew]. Our own analyses (Supplementary text, Supplementary Figs. 4-6) indicate, however that the impact of data recoding is is largely due to discarding information, not accommodating variation in amino acid composition. These findings indicate that recoding can be a problematic method for addressing heterogeneity.

@@

This dependence alone doesn't resolve the question of where the root falls in the animal tree of life, that depends on how we interpret this dependence. Focusing just on heterogeneous analyses of matrices with restricted outgroup sampling, should we believe the results with fewer equilibrium frequency categories or the results with many? XX

Overcoming the challenges associated with this difficult and important question goes beyond models. In particular, there are many genomic data that are highly relevant to this problem that have not yet been collected and analyzed in a phylogenetic context [@King:2017ie]. In particular, there are very few chromosome-level genome assemblies outside of Bilateria. There will be important advances in genome sequencing in coming years that will have a major impact on this problem. This is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and reduce the possibility of contamination. We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting difficult phylogenetic problem, and also to develop methods and tools that will help with other difficult phylogenetic problems.

@@

The data matrix was constructed using a semi-automated approach. Genes were translated into proteins, promiscuous domains were masked, all gene sequences from all species were compared to each other with blastp, genes were clustered based on this similarity with TribeMCL [@Enright:2002uq], and these clusters were filtered to remove those with poor taxon sampling and high rates of lineage-specific duplications. Gene trees were then constructed, and in clades of sequences all from the same species all but one sequence were removed (these groups are often due to assembly errors). The remaining gene trees with more than one sequence for any taxon were then manually inspected. If strongly supported deep nodes indicative of paralogy were found, the entire gene was discarded. If the duplications for a a small number of taxa were unresolved, all genes from those taxa were excluded. Genes were then realigned and sites were filtered with Gblocks [@Castresana:2000vy], resulting in a 77 taxon matrix. Some taxa in this matrix were quite unstable, which obscured other strongly-supported relationships. Unstable taxa were identified with leaf stability indices [@Thorley:1999kg], as implemented in phyutility [@Smith:2008gb], and removed from the matrix. This resulted in the 64-taxon matrix that is the focus of most of their analyses. Phylogenetic analyses were conducted under the Poisson+CAT model in PhyloBayes, and under the WAG model in MrBayes [@Ronquist:2003hx] and RAxML [@Stamatakis:2006wc].

@@

Note the main conclusion of this study is baed on the selection of genes that evolved slowly ribosomal RNA genes. The Ctenophora-sister was rejected based on the most saturated dataset and not supported by the matrix with the most closely related outgroups (Choanomalia matrix) [@nosenko2013deep].

#### Ryan *et al.* 2013
Ryan *et al.* sequenced the first ctenophore genome of *Mnemiopsis leidyi*. With the genome resources of *M. leidyi*, the authors constructed two phylogenomic datasets: a "Genome set" based on 13 animal genomes and a "EST Set" that also included 59 animals. They analyzed both matrices by site-homogeneous GTR+Gamma and site-heterogeneous Poisson+CAT models with three sets of outgroup sampling to evaluate the effect of outgroup selection to the ingroup topology for the Ryan2013_est matrix. The Orthologs were constructed based on the method of Hejnol *et al.* 2009. For the Ryan2013_genome matrix, they performed phylogenetic analyses with both gene content and sequence-baed analyses. Overall, their results strongly supported Ctenophora-sister in all datasets they analyzed using site-homogeneous model. The Poisson+CAT model of the genome dataset strongly supported of a clade of Ctenophora and Porifera as the sister group to all other Metazoa and Bayesian analysis on the EST dataset did not converge after 205 days (but strongly supported Porifera in Choaimalia matrix).

@@

#### Whelan et al. 2015
Whelan *et al.* 2015 [@whelan2015error] constructed a new phylogenomic dataset by eight new transcripomic data and investigated a range of possible sources of systematic error under multiple analyses (e.g. long-branch attraction, compositional bias, fast evolving genes, etc.). Putative orthologs were determined of each species using HaMStR using the model organism core ortholog set (same as Moroz *et al.* 2014) and subsequently removal of genes with too much missing data and potential paralogs.  The authors further filtered the full dataset to 24 sub-datasets by filtering genes with high long-branch scores; genes with high RSFV values; genes that are potential paralogs; fast evolving genes and progressively removal of outgroups. All the maximum likelihood analyses with site-homogeneous model and PartitionFinder strongly suggested Ctenophora-sister. CAT-GTR models only used in least saturated dataset 6 and 16 also strongly supported Ctenophora.

@@

Note that this data matrix has been extensively reanalyzed in many studies.

@@

Pisani *et al.* [@pisani2015genomic] reanalyzed representative datasets that supported Ctenophora-sister, including Ryan2013_est, Moroz2014_3d and Whelan2015 datasets. It was the first study showing that progressively removal of more distantly related outgroups could largely affect phylogenomic inference of the position of the root of animal phylogeny. The authors suggested that the inclusion of outgroups very distant from the ingroup can cause systematic errors due to long-branch attraction. Phylogenetic analyses were conducted under the Poisson+CAT and GTR models in PhyloBayes. They found GTR+CAT and Poisson+CAT models generally had better model-fit than site-homogeneous GTR models in these data matrice. Moreover, they found the support of Ctenophora-sister decreases when the exclusion of distantly related outgroups and the use of site-heterogeneous CAT models.

@@

Feuda *et al.* [@feuda2017improved] didn't generate any new data, instead they used the data-recoding methods to reanalyze two key datasets that support Ctenophora-sister (Whelan2015_D20, Chang2015 datasets). It was the first phylogenomic study that suggested recoding methods have better performance than non-recoding methods in the relationships at the root of the animal tree. The authors compared model adequacy using posterior predictive analyses from a set of site-homogeneous (WAG, LG, GTR, data-partitioning) and site-heterogeneous (CAT-GTR) models in non-recoding and recoding datasets. The results showed that data-recoding can significant reduce compositional heterogeneity in both datasets with CAT-GTR models and strongly supported Porifera-sister hypothesis (see more details in recoding section in  supplementary text).

@@

Whelan *et al.* [@whelan2016let] is the only study to evaluate performance of site-heterogeneous models and site-homogeneous model with data partitioning under the simulation framework. The simulation results suggested that Poisson+CAT model consistently performed worse than other models in simulation datasets. More importantly, the authors also showed that both Poisson+CAT and GTR+ CAT models could overestimated substitutional heterogeneity in many cases. They also reanalyzed datasets from Philippe 2009 and Nosenko 2013 using both CAT models and data partitioning with site-homogeneous model. The results indicated that Poisson + CAT model tends to recover less accurate trees and more importantly,  both GTR + CAT and data partitioning strongly supported Ctenophora-sister in reanalyses.

@@

Whelan *et al.* [@whelan2017ctenophore] added 27 new ctenophore transcriptomic data to explore animal-root position as well as relationships within Ctenophores. It significantly increased ctenophore taxon sampling than other studies. Putative orthologs were determined largely similar to Whelan2015, with the difference of a core Ctenophora core datasets were constructed here with more than 2000 genes. The subsequent filtering strategy was also similar to the previous study. All analyses using site-homogeneous and site-heterogeneous models strongly supported Ctenophora-sister hypothesis, even with CAT-GTR model in choanoazoa dataset. The main conclusions of this study were based on Whelan2017_full and Whelan2017_strict matrices.

@@

The final matrix was first analyzed using the Poisson+CAT model. Different that other phylobayes analyses, Simion *et al.* used a gene jackknife strategy based on 100 analyses to overcome the computational limitation because of the large data size. Each jackknife is baed on a random selection of ~ 25% of the genes. The Phylobayes with site-heterogeneous model strongly supported the Porifera-sister, whereas site-homogeneous strongly supported Ctenophora-sister in all datasets. Importantly, the authors compared the behavior of long-branch effect between site-heterogeneous and site-homogeneous models by progressively removal taxa and concluded higher sensitivity of site-homogeneous models to LBA than CAT models.

@@

> Our dataset outperforms previous metazoan gene superalignments in terms of data quality and quantity. Analyses with a best-fitting site-heterogeneous evolutionary model provide strong statistical support for placing sponges as the sister-group to all other metazoans, with ctenophores emerging as the second-earliest branching animal lineage.

>>>>

Yuanning Li: 0000-0002-2206-5804

Xing-Xing Shen: 0000-0002-8436-595X

Benjamin Evans:
Steven H.D. Haddock: 0000-0001-9420-4482

Antonis Rokas: 0000-0002-7248-6551

Casey W. Dunn: 0000-0003-0628-5150

@@

The models of sequence evolution in the studies considered here differ according to two primary components: the exchangeability matrix $E$ and amino acid equilibrium frequencies $\Pi$. The exchangeability matrix $E$ describes the relative rates at which one amino acid changes to another. The studies considered here use exchangeabilities that are the same between all amino acids (Poisson, also referred to as F81),or different; if different, the exchangeabilities can either be fixed based on previously empirically estimated rates (WAG or LG), or independently estimated from the data (GTR). The analyses considered here have site-homogeneous exchangeability models (site-homogeneous model), with the same matrix used for all sites in all genes. Or for data-partitioning analyses, some gene partitions use different site-homogeneous models. The equilibrium frequencies describe the expected frequency of each amino acid, which captures the fact that some amino acids are much more common than others. The analyses differ in whether they take a homogeneous approach and jointly estimate the same frequency across all sites in a partition, or add parameters that allow heterogeneous equilibrium frequencies that differ across sites. Heterogeneous approaches include CAT [@Lartillot:2004dq], which is implemented in the software PhyloBayes and has been widely applied to questions of deep animal phylogeny. The models that are applied in practice are heavily influenced by computational costs, model availability in software, and convention. While studies often discuss CAT and WAG models as if they are exclusive, we note that these particular terms apply to non-exclusive model components – CAT refers to heterogeneous equilibrium frequencies across sites and WAG to a particular exchangeability matrix. In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. To avoid confusion on this point, here we always specify the exchangeability matrix first (e.g., GTR), followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (e.g., CAT). If site homogeneous equilibrium frequencies are used, we refer to the exchangeability matrix alone. Gamma-rate heterogeneity, a scalar that accommodates total rate of change across sites, is used in almost every analysis conducted here and we generally omit its designation.

High-throughput sequencing allows investigators to readily assemble matrices with hundreds or thousands of protein-coding genes from a broad diversity of animal species (Extended Data Table 1). Studies of animal phylogeny have used a wide variety of different approaches to identifying and selecting genes and taxa for their matrices. As a result, the particular genes selected for analysis vary widely (Fig. 2B, Extended Data Fig. 1). Gene sampling varies in several ways, including in the fractions of single copy orthologs (e.g., BUSCO genes) and ribosomal protein genes in the matrix (Extended Data Fig. 2).

@@

**Fig. 2.** An overview of previous phylogenomic studieson animals. (A) A time-series showing the main conclusions of previous phylogenomic studies that indicated their analyses supported Ctenophora-sister (orange nodes) or Porifera-sister (green nodes) based on their main conclusion from each study. (B) Each of the primary matrices considered here, color coded by taxon sampling. Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled.

@@

**Fig. 3.**  Summary of phylogenomic results from previous studies and reanalysis conducted in this study. Each point represents a phylogenetic analysis with a support of bootstrap values less than `r bootstrap_threshold`% or a posterior probability less than `r posterior_prob_threshold`% are considered as unresolved. The clades are organized with increased outgroup sampling higher on the vertical axis, and the models are organized in general by increasing complexity in terms of numbers of parameters to the right on the horizontal axis. (A) Analyses transcribed from the literature (related to Supplementary Table 2). (B) New phylogenomic analyses conducted in this study (related to Supplementary Table 3).

@@

Our summary of previous phylogenetic analyses (Fig. 3A) shows that Ctenophora-sister is supported in analyses that span the full range of outgroup sampling and models used to date, including some analyses with restricted outgroup sampling and models that accommodate site-heterogeneous equilibrium frequencies with CAT. This is consistent with previous assessments of the problem [@whelan2016let], but is drawn from a much more extensive and systematic examination. The only analyses that support Porifera-sister have a reduced outgroup sampling (Choanimalia, Holozoa) and site heterogeneous models with CAT. Model adequacy assessments generally favor GTR+CAT over Poisson+CAT or site-homogeneous models [@pisani2015genomic; @feuda2017improved], but because GTR+CAT is so parameter rich, many analyses that use a model with GTR+CAT do not converge and thus renders the analysis untrustworthy / hard to interpret. The fact that Porifera-sister is recovered only for particular models with particular outgroup sampling indicates that model and outgroup interact, and that this interaction is fundamental to understanding the range of results obtained across analyses.

@@

One of the challenges of interpreting support for the placement of the animal root across published studies is that different programs, software versions, and settings have been used across analyses. This extensive variation makes it difficult to identify the primary factors that lead to different results. Here we first reanalyze the primary matrices from each study under the same conditions with IQ-TREE [@nguyen2014iq] with multiple evolutionary models. We selected this tool because it has greater model flexibility than most other phylogenetic tools [@zhou2017evaluating] (Fig. 3B; Supplementary Table 3). Importantly, it has C models (C10-C60 equilibrium frequencies) [@si2008empirical] that, like CAT, allow models to accommodate heterogeneity in equilibrium amino acid frequencies across sites.

@@

Among 15 previous studies,  three studies [@pisani2015genomic; @whelan2016let; @feuda2017improved]are based entirely on previously published data and gene-partition data is not available from one study [@Pick:2010eb]. For each of the published studies, we selected the matrix that was the primary focus of the manuscript for further analysis or has been reanalyzed extensively in other studies. For each of these matrices, we progressively trimmed taxon sampling to create Opisthokonta, Holozoa, and Choanimalia versions, where permitted by original outgroup sampling. This produced `r analyses_new$matrix%>% unique()%>% length()`data matrices from 11 studies.

@@

Once we established this large difference in category number, we sought to test its direct impact on animal root position. It is not possible to use more than 60 categories for C models in IQ-TREE, but the number of categories can be set *a priori* in Phylobayes using nCAT. We therefore varied the number of categories in Phylobayes analyses (Fig. 4; Supplementary Table 6). We found that Poisson+nCAT60 analyses in PhyloBayes, like Poisson+C60 and WAG+C60 IQ-TREE analyses, provide strong support of Ctenophora-sister. This indicates that the difference in results between unconstrained CAT analyses in PhyloBayes and C60 analyses in IQ-TREE is not due to differences in the software, but due to the large difference, in excess of ten-fold, in the number of site categories. When we varied the number of categories in PhyloBayes, we observe the transition from support for Ctenophora-sister to an unresolved root to Porifera-sister (Fig. 4). For example, for the Whelan2017_strict matrix this transition occurs between 60 -- 120 categories when using Poisson-CAT model.

@@

These results further clarify when analyses support Porifera-sister (e.g., Philippe2009, Whelan2017). Outgroup sampling must be restricted (Choanimalia), a very large number of site categories must be used (unconstrained CAT, giving hundreds of categories), and a simplistic Poisson rather than GTR exchange matrix must be used. Analyses under other conditions either support Ctenophora-sister or are unresolved.

@@

ggarrange(p1, p2, p3, p4,  ncol  = 2, nrow = 2, labels = c("A", "B", "C", "D"))

@@

Although a higher proportion of genes support Ctenophora-sister with site-heterogeneous C60 models, the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models. In an extreme case, in matrices from Ryan2013_est nearly 30% of genes changed from strong to weak Ctenophora-sister signal ($\Delta$|lnL|<2) (Extended Data Fig. 3; Supplementary Table 7). In contrast to the C60 models, the phylogenetic signal largely increased in Poisson+CAT models in PhyloBayes towards Porifera-sister, and the outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig. 3). For example in Whelan2017_full matrix, we found that the number of genes that support Ctenophora-sister in analyses with CAT decreases from 57.5% in matrices with distant outgroups (Holozoa) to 35.4% when outgroups are restricted (Choanimalia, Supplementary Table 7).

@@

Other approaches have been taken to addressing base compositional heterogeneity across taxa. For example, Feuda *et al.* [@Feuda:2017ew] recoded the full set of twenty amino acids into six groups of amino acids. These groups tend to have more frequent evolutionary changes within them than between them [@Susko:2007ds]. Recoding could, like CAT and C models, address variation across sites, but it could also accommodate variation across lineages, and it was suggested that this approach favors Porifera-sister [@Feuda:2017ew]. However, Our own analyses (Supplementary text, Supplementary Figs. 4-6) by comparing performance of random and empirical recoding indicate, however that the impact of data recoding is largely due to discarding information, not accommodating variation in amino acid composition. Consistent with a recent simulation study on data recoding [@hernandez2019six], these findings indicate that recoding can be a problematic method for addressing heterogeneity.

@@

This dependence alone doesn't resolve the question of where the root falls in the animal tree of life, that depends on how we interpret this dependence. Focusing just on heterogeneous analyses of matrices with restricted outgroup sampling, should we believe the results with fewer equilibrium frequency categories or the results with many? Moreover, taxa sampling also play an important role in this question since data matrices with more ctenophores (and outgroups) generally results in more support for Ctenophora-sister.

Overcoming the challenges associated with this difficult and important question goes beyond models. In particular, there are many genomic data that are highly relevant to this problem that have not yet been collected and analyzed in a phylogenetic context [@King:2017ie]. In particular, there are very few chromosome-level genome assemblies outside of Bilateria. There will be important advances in genome sequencing in coming years that will have a major impact on this problem. Instead of continuing to use previous matrices that has limited taxa sampling (e.g., Chang2015 and Whelan2015), without analyses with greater taxon sampling, confound conclusions as greater taxon sampling should result in better tree inference. [XX Nathan: But, I think there needs to be a call for generating datasets with the most available non-bilaterian taxon sampling. Continuing to use datasets from Chang and Whelan, without analyses with greater taxon sampling, confound conclusions as greater taxon sampling should result in better tree inference.] This is about more than more data - high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction and reduce the possibility of contamination. We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting difficult phylogenetic problem, and also to develop methods and tools that will help with other difficult phylogenetic problems.

@@

The data matrix was constructed using a semi-automated approach. Genes were translated into proteins, promiscuous domains were masked, all gene sequences from all species were compared to each other with blastp, genes were clustered based on this similarity with TribeMCL [@Enright:2002uq], and these clusters were filtered to remove those with poor taxon sampling and high rates of lineage-specific duplications. Gene trees were then constructed, and in clades of sequences all from the same species all but one sequence were removed (these groups are often due to assembly errors). The remaining gene trees with more than one sequence for any taxon were then manually inspected. If strongly supported deep nodes indicative of paralogy were found, the entire gene was discarded. If the duplications for a small number of taxa were unresolved, all genes from those taxa were excluded. Genes were then realigned and sites were filtered with Gblocks [@Castresana:2000vy], resulting in a 77 taxon matrix. Some taxa in this matrix were quite unstable, which obscured other strongly-supported relationships. Unstable taxa were identified with leaf stability indices [@Thorley:1999kg], as implemented in phyutility [@Smith:2008gb], and removed from the matrix. This resulted in the 64-taxon matrix that is the focus of most of their analyses. Phylogenetic analyses were conducted under the Poisson+CAT model in PhyloBayes, and under the WAG model in MrBayes [@Ronquist:2003hx] and RAxML [@Stamatakis:2006wc].

@@

#### Ryan *et al.* 2013
Ryan *et al.* sequenced the first ctenophore genome of *Mnemiopsis leidyi*. With the genome resources of *M. leidyi*, the authors constructed two phylogenomic datasets: a "Genome set" based on 13 animal genomes and a "EST Set" that also included 59 animals. They analyzed both matrices by site-homogeneous GTR+Gamma and site-heterogeneous Poisson+CAT models with three sets of outgroup sampling to evaluate the effect of outgroup selection to the ingroup topology for the Ryan2013_est matrix. The Orthologs were constructed based on the method of Hejnol *et al.* 2009. For the Ryan2013_genome matrix, they performed phylogenetic analyses with both gene content and sequence-based analyses. Overall, their results strongly supported Ctenophora-sister in all datasets they analyzed using site-homogeneous model. The Poisson+CAT model of the genome dataset strongly supported of a clade of Ctenophora and Porifera as the sister group to all other Metazoa and Bayesian analysis on the EST dataset did not converge after 205 days (but strongly supported Porifera in Choaimalia matrix).

@@

#### Whelan et al. 2015
Whelan *et al.* 2015 [@whelan2015error] constructed a new phylogenomic data matrix with eight new transcripomic data and investigated a range of possible sources of systematic error under multiple analyses (e.g. long-branch attraction, compositional bias, fast evolving genes, etc.). Putative orthologs were determined of each species using HaMStR using the model organism core ortholog set (same as Moroz *et al.* 2014) and subsequently removal of genes with too much missing data and potential paralogs.  The authors further filtered the full dataset to 24 sub-datasets by filtering genes with high long-branch scores; genes with high RSFV values; genes that are potential paralogs; fast evolving genes and progressively removal of outgroups. All the maximum likelihood analyses with site-homogeneous model and PartitionFinder strongly suggested Ctenophora-sister. GTR-CAT models only used in slow-evolving data matrices 6 and 16 also strongly supported Ctenophora.

@@

Note that this data matrix has been extensively reanalyzed in many studies although there is very little ctenophore sampling and Myxozoans contain extremely long branch in the tree.

@@

Pisani *et al.* [@pisani2015genomic] reanalyzed representative datasets that supported Ctenophora-sister, including Ryan2013_est, Moroz2014_3d and Whelan2015 datasets. It was the first study showing that progressively removal of more distantly related outgroups could largely affect phylogenomic inference of the position of the root of animal phylogeny. The authors suggested that the inclusion of outgroups very distant from the ingroup can cause systematic errors due to long-branch attraction. Phylogenetic analyses were conducted under the Poisson+CAT and GTR models in PhyloBayes. They found Poisson+CAT models generally had better model-fit than site-homogeneous GTR models in these data matrice. Moreover, they found the support of Ctenophora-sister decreases when the exclusion of distantly related outgroups and the use of site-heterogeneous CAT models.

@@

Feuda *et al.* [@feuda2017improved] didn't generate any new data, instead they used the data-recoding methods to reanalyze two key datasets that support Ctenophora-sister (Whelan2015_D20, Chang2015 datasets). It was the first phylogenomic study that suggested recoding methods have better performance than non-recoding methods based on recovering Porifera-sister hypothesis. The authors compared model adequacy using posterior predictive analyses from a set of site-homogeneous (WAG, LG, GTR, data-partitioning) and site-heterogeneous (CAT-GTR) models in non-recoding and recoding datasets. The results showed that data-recoding can significant reduce compositional heterogeneity in both datasets with CAT-GTR models and strongly supported Porifera-sister hypothesis (see more details in recoding section in  supplementary text).

@@

Whelan *et al.* [@whelan2016let] is the only study to evaluate performance of site-heterogeneous models and site-homogeneous model with data partitioning under the simulation framework. The simulation results suggested that Poisson+CAT model consistently performed worse than other models in simulation datasets. More importantly, the authors also showed that both Poisson+CAT and GTR+ CAT models could overestimated substitutional heterogeneity in almost every cases. They also reanalyzed datasets from Philippe 2009 and Nosenko 2013 using both CAT models and data partitioning with site-homogeneous model. The results indicated that Poisson + CAT model tends to recover less accurate trees and both GTR + CAT and data partitioning strongly supported Ctenophora-sister in reanalyses.

@@

Whelan *et al.* [@whelan2017ctenophore] added 27 new ctenophore transcriptomic data to explore animal-root position as well as relationships within Ctenophores. It significantly increased ctenophore taxon sampling than other studies. Putative orthologs were determined largely similar to Whelan2015. The subsequent filtering strategy was also similar to the previous study. All analyses using site-homogeneous and site-heterogeneous models strongly supported Ctenophora-sister hypothesis, even with CAT-GTR model in choanoazoa dataset. The main conclusions of this study were based on Whelan2017_full and Whelan2017_strict matrices.

@@

The final matrix was first analyzed using the Poisson+CAT model. Different that other phylobayes analyses, Simion *et al.* used a gene jackknife strategy based on 100 analyses to overcome the computational limitation because of the large data size. Each jackknife is based on a random selection of ~ 25% of the genes. The Phylobayes with site-heterogeneous model strongly supported the Porifera-sister, whereas site-homogeneous strongly supported Ctenophora-sister in all datasets. Importantly, the authors compared the behavior of long-branch effect between site-heterogeneous and site-homogeneous models by progressively removal taxa and concluded higher sensitivity of site-homogeneous models to LBA than CAT models.

@@

> Our dataset outperforms previous metazoan gene super alignments in terms of data quality and quantity. Analyses with a best-fitting site-heterogeneous evolutionary model provide strong statistical support for placing sponges as the sister-group to all other metazoans, with ctenophores emerging as the second-earliest branching animal lineage.
